# Getting Started

## Introduction
R based programs to analyze and visualize embryo lineage after the cell data are extracted from 3D movies in Acetree
Intended to process data tables where: 

* each table represents one embryo
* each row represents one cell at a specific time point, with columns: 
  + `time`: imaging time this cell read is created
  + `cell`: cell name
  + `x`: x position
  + `y`
  + `z`
  + `blot`: reporter expression

`LineageProcessing.R` contains functions to process one single embryo entity, and slices from embryo entities, which are the dependencies of other codes 
`CD_Processing.R` contains functions to work with a directory of several embryo data tables. Including functions to load data tables, retrieve specific cells and lineages data from multiple embryos, and operations that are only possible with multiple embryos (e.g. depth correction)
`drawEmb.R` contains functions that plot all nuclei in an embryo in 3D

## Before you start
Before running the examples below, please ensure that you have installed [LIVEtools](https://github.com/johnmurraylab/LIVE_tools#installation) and its dependencies. 
```{r}
library("LIVEtools")
```

## Understanding Data Structures

### Single Embryo Data
When loading a single embryo file with `readEmbryoTable()`, the function returns a **list** containing:
1. `[[1]]` - The embryo data.frame (cell tracking table)
2. `[[2]]` - A boolean indicating if time correction was applied

**Important**: To access the actual embryo data.frame, you must use `[[1]]`:
```{r}
# Returns a list
embryo_list <- readEmbryoTable(file = "sample_data/20241108_JIM767_25_L5-edit.zip")

# Extract the data.frame - THIS IS REQUIRED for downstream functions
embryo <- embryo_list[[1]]

# Attempting to use embryo_list directly in plotting functions will cause errors
# because they expect a data.frame, not a list
```

### Multi-Embryo Data Structure
When processing multiple embryos with `CD_In()`, the input directory should contain:
```
data_directory/
  ├── CDembryo1.csv       # Cell tracking data table for embryo 1
  ├── CDembryo2.csv       # Cell tracking data table for embryo 2
  ├── ...
  ├── TIMEembryo1.csv     # Actual imaging times for time correction (optional)
  ├── TIMEembryo2.csv
  ├── ...
  ├── embryo1AuxInfo.csv  # Metadata (optional): coordinate scales, rotation status, etc. 
  ├── embryo2AuxInfo.csv
  └── ...
```

`CD_In()` returns a nested list structure:
```
embryos                   # Main list object
  ├── $CD                 # List of cell tracking data.frames
  |   ├── $embryo1        # data.frame from CDembryo1.csv (time-corrected if TIME file exists)
  |   ├── $embryo2        # data.frame from CDembryo2.csv
  |   └── ...
  └── $info               # List of metadata data.frames
      ├── $embryo1        # data.frame from embryo1AuxInfo.csv + processing flags
      ├── $embryo2        # data.frame from embryo2AuxInfo.csv + processing flags
      └── ...
```

The `$info` data.frames contain:
- `name`: Embryo identifier
- `depth_corrected`: Boolean flag
- `time_corrected`: Boolean flag  
- `rotated`: Boolean flag
- Additional custom metadata from AuxInfo files

### Accessing Individual Embryos

**Method 1: Extract by index (quick but loses identity)**
```{r}
# Extract first embryo's data
exEmbryo <- embryos[["CD"]][[1]]

# WARNING: This loses the embryo name and metadata!
# You won't know which embryo this is without additional tracking
```

**Method 2: Extract with metadata (recommended)**
```{r}
# Extract both data and metadata together
exEmbryo <- embryos[["CD"]][[1]] 
exEmbryo_info <- embryos[["info"]][[1]]

# Now you can access identifying information:
embryo_name <- exEmbryo_info$name
embryo_name  # Returns the embryo identifier

# Or extract by name (if you know it):
exEmbryo <- embryos[["CD"]][["embryo1"]]
exEmbryo_info <- embryos[["info"]][["embryo1"]]
```

**Method 3: Extract by name from the start (recommended)**
```{r}
# If you know the embryo name, access it directly:
embryo_name <- "embryo1"
exEmbryo <- embryos[["CD"]][[embryo_name]]
exEmbryo_info <- embryos[["info"]][[embryo_name]]
```

### Working with Multi-Embryo Data

Many functions in this package are designed to work with the entire `embryos` list structure. For example, `plotBlotLine()` automatically processes all embryos and plots them together. However, 3D visualization and tree plotting functions (`drawEmbLine()`, `drawEmbVal()`, `CD_tree_plot()`) require a **single embryo data.frame** as input.

To iterate over all embryos for individual processing:
```{r}
# Get list of embryo names
embryo_names <- names(embryos[["CD"]])

# Process each embryo
for (emb_name in embryo_names) {
  this_embryo <- embryos[["CD"]][[emb_name]]
  this_info <- embryos[["info"]][[emb_name]]
  # Your processing code here
  # e.g., create 3D plots, save with embryo-specific file name
}
```

**Note on Advanced Multi-Embryo Operations**: 
- **Averaged trees**: Not currently implemented, but could be achieved by aggregating data across embryos using `aggEmbryos()` with custom post-processing
- **Averaged 3D positions**: Possible by aligning embryos to equivalent developmental stages (using `alignTime()`), then averaging x/y/z coordinates across embryos for each cell type at matched timepoints

## Example Workflow
### Process a Single Embryo
```{r}
# Load a single embryo file (from StarryNite format zip file)
# Genotype: JIM767 is tbx-35(tm1789) II; pha-4::GFP; ...
embryo_list <- readEmbryoTable(file = "sample_data/20241108_JIM767_25_L5-edit.zip")

# IMPORTANT: Extract the data.frame from the returned list using [[1]]
# The list contains: [[1]] = data.frame, [[2]] = boolean for time correction
embryo <- embryo_list[[1]]

# Verify we have a data.frame
class(embryo)  # Should print "data.frame"
```
#### Plot the embryo nucleus positions in 3D. 
Plot the lineages
```{r}
time = 139
figO <- drawEmbLine(
  #no lineage argument provided, so the function will group cells into ABa, ABp, MS, E, C, D, P4
  embryo, time, cellSize=10, title = "Default 7 Lineages", 
  xSize=0.08651, ySize=0.08651, zSize=0.5 #unprocessed data can have different x,y,z scales, specify here
  )
figO[[1]]
```
#### Align embryo position/rotation and then plot
`totalRePosition` will trsnslate all embryo cells to center the embryo at (0,0,0), align embryo anterior with x+ direction, left with y+, dorsal with z+.
```{r}
rotatedEmbryo <- totalRePosition(
  embryo, time = time, indicatorP = "C", indicatorD = "Cxa", indicatorV = "MSxxp",
  xSize=0.08651, ySize=0.08651, zSize=0.5) #the xyz dimension is corrected during rotation
#so we no longer have to provide xyz dimension size in plotting function (defualt is 1)
figRotated <- drawEmbLine(
  rotatedEmbryo, time, title = "Default Lineages Aligned", cellSize=10
  )
figRotated[[1]]
```
#### Customize Lineage 3D plot
`drawEmbLine` function allows user to select an array of lineages to highlight, then assign color and alpha (transparency) for **each selected lineage** as two array with corresponding ordering. 

- Nucleus not selected will be grey and transparent. 
- The colors and alphas should be delivered as lists where each element correspond to one lineage in "lineges" list. 
- If no colors are indicated, or the length of colors array doesn't match that of lineages list, the function will automatically assign unique colors from viridis palette.
```{r}
figABp <- drawEmbLine(
  rotatedEmbryo, 
  lineages = c("ABp", "C"), 
  colors = c(rgb(0.7,0,0.5),rgb(0.3,1,0)), #"ABp" will have rgb(0.7,0,0.5),
  alphas = c(0.3, 1), 
  cellSize=6, time = time
  )
figABp[[1]]
```
#### Plot 3D with marker gene expression
The default color scale is Viridis
```{r}
figVal <- drawEmbVal(
  rotatedEmbryo, time = time, title = "expression in default color scale",
  cellSize=10)
figVal[[1]]
```
#### Gene expression with customized color scale
`drawEmbVal` function offers several options to customize colors through `colorScheme` argument

- One may supply a list of vectors, where each vector is a mapping from a normalized value (0 for minimum, 1 for maximum, 0.5 for halfway in the range) to an R color representation. The mapping for 0 and 1 are mandatory. 
- The `colorScheme` argument can also be a palette name string of the following list: Blackbody, Bluered, Blues, Cividis, Earth, Electric, Greens, Greys, Hot, Jet, Picnic, Portland, Rainbow, RdBu, Reds, Viridis, YlGnBu, YlOrRd.

```{r}
figVal <- drawEmbVal(
  rotatedEmbryo, time = time, title = "expression with black-blue-white-red color scale",
  # here, define pure balck for minimum value, pure blue for 1/3 between maximum and minimum etc.
  colorScheme = list(c(0,rgb(0,0,0)), c(1/3,rgb(0,0,1)), c(2/3,rgb(1,1,1)), c(1,rgb(1,0,0))),
)
figVal[[1]]
```
#### Lineage Tree colored by Expression
The expression color scaling methods can be delivered as `transform` argument. 
- the default "identity" makes a linear mapping between expression and color scale
- "log2", "log10", "log" will bring resolute lower expression values through a log color scale. 
```{r}
treeplot <- CD_tree_plot(
  rotatedEmbryo, 
  min_gain=-1000, max_gain = 8000, transform = "identity", 
  root = "ABara", end_time=time)
treeplot
```

### Process a Batch of Embryos
#### Read a directory of embryo tables. 
```{r} 
embryos <- CD_In(directory = "sample_data/JIM767_25/", prefix = "CD",
                 time.prefix = "TIME", time.suffix = ".csv",
                 info.prefix = "", info.suffix = "AuxInfo.csv")
```

#### Optional: implement depth correction.
```{r}
# get the correction parameters for the model of fluoresence loss by depth 
model <- depthCorrectionParm(embryos, lineage = c("ABara", "ABalp", "E"), alignCell = "E", startT = 30, endT = 150, exc_zMin = 0.2, exc_zMax = 1, zMax = 67)
# correct the embryos' data with the model
embryos <- embryos |> 
  dataCorrection(lineage=c("ABara", "ABalp", "P1"), model = model, zMax = 67, exc_zMin = 0.2, exc_zMax = 1)
```
#### Plot the reporter expression against time
`plotBlotLine` function plot the expression of selected cells, averaged in each embryo at each time point. Each embryo will be represented by a unique curve.
- Synchronize start point for each embryo: if an `aligningCell` argument is delivered, the function will translate the time points in each embryo to make the last time point of `aligningCell` lifespan equal to `align_t` time. 
```{r}
yrange = list(0,16000) #default figure x,y range
xrange = list(-50,150)
embNames <- names(embryos[["CD"]]) 
emb_palette <- viridis::viridis(length(embNames)) |> setNames(embNames)#set up palette to make color consistent for the same embryo across different plots
cellsExp <- strsplit("ABal, ABalp, ABalpa, ABalpaa, ABalpaaa, ABalpaaaa, ABalpaaaaa", split=", ")[[1]]
fig_blot_cells <- embryos|>plotBlotLine(
  title = 'specific cells example', color_palette = emb_palette, 
  aligningCell = "ABal", align_t = 0,
  cells = cellsExp, 
  xrange = xrange, yrange = yrange)
fig_blot_cells
```
#### Track Expression with Various Cell and Lineage Representations
- You can take a cell and all its daughter cells into account by putting it into `lineages` argument
- Both `cells` and `lineages` accepts "x" within cell name, so "ABpxpappp" would include both "ABplpappp" and "ABprpappp"
- The function averages all cells selected at each time point, the user can mix `cells` and `lineages` from with different parents. 
```{r}
fig_blot_lines <- embryos|>plotBlotLine(title = 'lineages example',
                            aligningCell = "ABalp", align_t = 0, 
                            cells = c("Exxx", "Exxxx"), #every 3rd and 4th generation E cell
                            lineages = c("ABalp","ABara"), 
                            color_palette = emb_palette,
                            xrange = xrange, yrange = yrange)
fig_blot_lines
```
#### Extract a Single Embryo from Multi-Embryo Data for 3D Visualization

When you need to visualize individual embryos from a multi-embryo dataset, extract both the data and metadata to preserve embryo identity:
```{r}
library(reticulate)
reticulate::use_virtualenv("../LIVEtools_venv/")
# Extract first embryo with its metadata
embryo_names <- names(embryos[["CD"]])
first_emb_name <- embryo_names[1]

exEmbryo <- embryos[["CD"]][[first_emb_name]]
exEmbryo_info <- embryos[["info"]][[first_emb_name]]

# Now we can identify this embryo
message(paste("Processing embryo:", exEmbryo_info$name))

# Rotate and visualize
exEmbryo <- totalRePosition(
  exEmbryo, time = time, indicatorP = "C", indicatorD = "Cxa", indicatorV = "MSxxp",
  xSize=0.08651, ySize=0.08651, zSize=0.5)

fig_3D_list <- drawEmbLine(exEmbryo, time)
fig_3D_list[[1]]
```

####save one frame as a png, keeping color scale legend
```{r}
saveEmbImg(fig_3D_list[[1]], output_file="example_3D.png",keepLabels=TRUE)
```

##save a collection of frames that could be combined make a movie (in FIJI or other image analysis software)
```{r}
outdir <- "out/"
dir.create("out", recursive = TRUE, showWarnings = FALSE)

for(time in 1:150){
    if(time %% 10 == 0){
        message(time)
    }
    fig_3D_list <- drawEmbLine(
        exEmbryo, time, cellSize=15,
        xSize=1, ySize=1, zSize=1 #unprocessed data can have different x,y,z scales, specify here
    )
    saveEmbImg(fig_3D_list[[1]], output_file=paste0(outdir, "example_3D", time,".png"),keepLabels=TRUE, width = 800, height = 650, zoom=1.3)
}
```
