# Getting Started

## Introduction
R based programs to analyze and visualize embryo lineage after the cell data are extracted from 3D movies in Acetree
Intended to process data tables where: 

* each table represent one embryo
* each row represents one cell at a specific time point, with columns: 
  + `time`: imaging time this cell read is created
  + `cell`: cell name
  + `x`: x position
  + `y`
  + `z`
  + `blot`: reproter expression

`LineageProcessing.R` contains functions to process one single embryo entity, and slices from embryo entities, which are the dependencies of other codes. INcluding functions 
`CD_Processing.R` are functions to work with a directory of several embryo data tables. Including functions to load data tables, retrieve specific cells and lineages data from multiple embryos, and operations that are only possible with multiple embryos (e.g. depth correction)
`drawEmb.R` contains functions that plot all nucleus in an embryo in 3D
```{r}
devtools::load_all("../LinAlign/")
```
## Example Workflow
### Process a Single Embryo
```{r}
#load a embryo file
embryo <- readEmbryoTable(file = "sample_data/20241108_JIM767_25_L5-edit.zip")[[1]]
```
#### Plot the embryo nucleus positions in 3D. 
Plot the lineages
```{r}
time = 139
figO <- drawEmbLine(
  embryo, 139, 
  xSize=0.08651, ySize=0.08651, zSize=0.5 #unprocessed data can have different x,y,z scales, specify here
  )
figO[[1]]
```
#### Optional: Align embryo position/rotation and then plot
```{r}
rotatedEmbryo <- totalRePosition(
  embryo, time = 200, indicatorP = "C", indicatorD = "Cxa", indicatorV = "MSxxp",
  xSize=0.08651, ySize=0.08651, zSize=0.5) #the xyz dimension is corrected during rotation
figRotated <- drawEmbLine(
  rotatedEmbryo, 200, xSize = 1, ySize = 1, zSize = 1
  )
figRotated[[1]]
```
#### Plot 3D with marker gene expression
```{r}
figExp <- drawEmbVal(rotatedEmbryo, time = 139)
figExp[[1]]
```
#### Lineage Tree colored by Expression
```{r}
treeplot <- CD_tree_plot(
  rotatedEmbryo, min_gain=-1000, max_gain=16000, transform = "identity", root = "MSaa")
treeplot
```

### Process a Batch of Embryos
#### Read a directory of embryo tables. 
```{r} 
embryos <- CD_In(directory = "sample_data/JIM767_25/", prefix = "CD",
                 time.prefix = "TIME", time.suffix = ".csv",
                 info.prefix = "", info.suffix = "AuxInfo.csv")
```

#### Optional: implement depth correction.
```{r}
# get the correction parameters for the model of fluoresence loss by depth 
model <- depthCorrectionParm(embryos, lineage = c("ABara", "ABalp", "E"), alignCell = "E", startT = 30, endT = 150, exc_zMin = 0.2, exc_zMax = 1, zMax = 67)
# correct the embryos' data with the model
embryos <- embryos |> 
  dataCorrection(lineage=c("ABara", "ABalp", "P1"), model = model, zMax = 67, exc_zMin = 0.2, exc_zMax = 1)
```
#### Plot the reporter expression against time for a particular lineage or specified cells for all embryos.
```{r}
yrange = list(0,16000)
xrange = list(-50,150)
# several specified cells in each embryo
embNames <- names(embryos[["CD"]]) #set up palette to make color consistent for the same embryo across different plots
library(viridis)
emb_palette <- turbo(length(embNames)) |> setNames(embNames)
cellsExp <- strsplit("ABal, ABalp, ABalpa, ABalpaa, ABalpaaa, ABalpaaaa, ABalpaaaaa", split=", ")[[1]]
fig_blot_cells <- embryos|>plotBlotLine(title = 'specific cells example', 
                            aligningCell = cellsExp[1], align_t = 0, cells = cellsExp, 
                            color_palette = emb_palette,
                            xrange = xrange, yrange = yrange)
fig_blot_cells
```
```{r}
# several lineages in each embryo
lineExp <- c("ABalp", "ABara", "ABplpappp", "ABprpappp")
fig_blot_lines <- embryos|>plotBlotLine(title = 'lineages example', 
                            aligningCell = lineExp[1], align_t = 0, lineages = lineExp, 
                            color_palette = emb_palette,
                            xrange = xrange, yrange = yrange)
fig_blot_lines
```
