# Getting Started

## Introduction
R based programs to analyze and visualize embryo lineage after the cell data are extracted from 3D movies in Acetree
Intended to process data tables where: 

* each table represents one embryo
* each row represents one cell at a specific time point, with columns: 
  + `time`: imaging time this cell read is created
  + `cell`: cell name
  + `x`: x position
  + `y`
  + `z`
  + `blot`: reporter expression

`LineageProcessing.R` contains functions to process one single embryo entity, and slices from embryo entities, which are the dependencies of other codes 
`CD_Processing.R` contains functions to work with a directory of several embryo data tables. Including functions to load data tables, retrieve specific cells and lineages data from multiple embryos, and operations that are only possible with multiple embryos (e.g. depth correction)
`drawEmb.R` contains functions that plot all nuclei in an embryo in 3D

## Installation
Before running the examples below, please ensure the package and its dependencies are installed.

### 1. Install Dependencies

This package depends on several CRAN and Bioconductor packages.  
You can install all dependencies automatically using the following commands:

#### Install CRAN dependencies
```{r}
cran_deps <- c(
  "data.table", "dplyr", "plotly", "reticulate", "ggplot2", "tidyr", "viridis" 
) # <-- replace with actual CRAN dependencies from DESCRIPTION
cran_missing <- setdiff(cran_deps, rownames(installed.packages()))
if (length(cran_missing)) {
  install.packages(cran_missing)
}
```
#### Install Bioconductor dependencies
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager") #first, make sure Bioconductor package manager is present

bioc_deps <- c(
  "ggtree" #Bioconductor packages to install
)
bioc_missing <- setdiff(bioc_deps, rownames(installed.packages())) #check missing and install
if (length(bioc_missing)) {
  BiocManager::install(bioc_missing)
}
```
### Install LIVEtools
Download LIVEtools release package (with file name `LIVEtools_?.?.?.????.tar.gz` where the `?`s are version numbers) from [**Releases**]{https://github.com/johnmurraylab/LIVE_tools/releases} on github repository

Then run the following command
```{r}
#replace package name with the actual release file name you downloaded
install.packages("full/path/to/tar.gz/file", repos = NULL, type = "source")
```

##or use this:
```{r}
devtools::install_github("johnmurraylab/LIVE_tools")
library(LIVEtools)
```

## Example Workflow
### Process a Single Embryo

```{r}
#load a embryo file (from starrynight format zip file)
#genotype of JIM767 is tbx-35(tm1789) II; pha-4::GFP; irEx274[ZK177 + unc-119::CFP + rol-6(su1006)]; stIs10226[his-72p::HIS-24::mCherry::let-858 3' UTR + unc-119] OR stIs10116 [his-72(promoter)::his-24::mCherry::let-858 3'UTR + unc-119(+)]; ltIs37[pie-1p::mCherry::his-58 + unc-119] IV; 
embryo <- readEmbryoTable(file = "sample_data/20241108_JIM767_25_L5-edit.zip")[[1]]
```
#### Plot the embryo nucleus positions in 3D. 
Plot the lineages
```{r}
time = 139
figO <- drawEmbLine(
  embryo, time, cellSize=10, 
  xSize=0.08651, ySize=0.08651, zSize=0.5 #unprocessed data can have different x,y,z scales, specify here
  )
figO[[1]]
```
#### Align embryo position/rotation and then plot
```{r}
rotatedEmbryo <- totalRePosition(
  embryo, time = time, indicatorP = "C", indicatorD = "Cxa", indicatorV = "MSxxp",
  xSize=0.08651, ySize=0.08651, zSize=0.5) #the xyz dimension is corrected during rotation


figRotated <- drawEmbLine(
  rotatedEmbryo, time, xSize = 1, ySize = 1, zSize = 1, cellSize=10
  )
figRotated[[1]]
```
#### Highlight a Selective Set of Sells with Customized Colors and Display Options
`drawEmbLine` function allows user to select a list of lineages to highlight, then assign color and alpha (transparency) for each selected lineage. Nucleus not selected will be grey and transparent. The colors and alphas should be delivered as lists where each element correspond to one lineage in "lineges" list. If no colors are indicated, or the length of colors list doesn't match that of lineages list, the function will automatically assign unique colors from Viridis palette.
```{r}
figABp <- drawEmbLine(
  rotatedEmbryo, 
  lineages = c("ABp", "C"), colors = c(rgb(0.7,0,0.5),rgb(0.3,1,0)), alphas = c(0.3, 1), cellSize=6, 
  time, xSize = 1, ySize = 1, zSize = 1
  )
figABp[[1]]
```
#### Plot 3D with marker gene expression
The default color scale is Viridis
```{r}
figVal <- drawEmbVal(
  rotatedEmbryo, time = time,
  cellSize=10)
figVal[[1]]
```
#### Gene expression with customized color scale
`drawEmbVal` function offers several options to customize colors through `colorScheme` argument

- One may supply a list of vectors, where each vector is a mapping from a normalized value (0 for minimum, 1 for maximum, 0.5 for halfway in the range) to an R color representation. The mapping for 0 and 1 are mandatory. 

- The `colorScheme` argument can also be a palette name string of the following list: Blackbody, Bluered, Blues, Cividis, Earth, Electric, Greens, Greys, Hot, Jet, Picnic, Portland, Rainbow, RdBu, Reds, Viridis, YlGnBu, YlOrRd.

```{r}
figVal <- drawEmbVal(
  rotatedEmbryo, time = time, 
  colorScheme = list(c(0,rgb(0,0,0)), c(1/3,rgb(0,0,1)), c(2/3,rgb(1,1,1)), c(1,rgb(1,0,0))),
  cellSize=10)
figVal[[1]]
```
#### Lineage Tree colored by Expression

```{r}
treeplot <- CD_tree_plot(
  rotatedEmbryo, min_gain=-1000, max_gain=16000, transform = "identity", root = "MS", end_time=time)
treeplot
```

### Process a Batch of Embryos
#### Read a directory of embryo tables. 
```{r} 
embryos <- CD_In(directory = "sample_data/JIM767_25/", prefix = "CD",
                 time.prefix = "TIME", time.suffix = ".csv",
                 info.prefix = "", info.suffix = "AuxInfo.csv")
```

#### Optional: implement depth correction.
```{r}
# get the correction parameters for the model of fluoresence loss by depth 
model <- depthCorrectionParm(embryos, lineage = c("ABara", "ABalp", "E"), alignCell = "E", startT = 30, endT = 150, exc_zMin = 0.2, exc_zMax = 1, zMax = 67)
# correct the embryos' data with the model
embryos <- embryos |> 
  dataCorrection(lineage=c("ABara", "ABalp", "P1"), model = model, zMax = 67, exc_zMin = 0.2, exc_zMax = 1)
```
#### Plot the reporter expression against time for a particular lineage or specified cells for all embryos.
```{r}
yrange = list(0,16000)
xrange = list(-50,150)
# several specified cells in each embryo
embNames <- names(embryos[["CD"]]) 
emb_palette <- viridis::viridis(length(embNames)) |> setNames(embNames)#set up palette to make color consistent for the same embryo across different plots
cellsExp <- strsplit("ABal, ABalp, ABalpa, ABalpaa, ABalpaaa, ABalpaaaa, ABalpaaaaa", split=", ")[[1]]
fig_blot_cells <- embryos|>plotBlotLine(title = 'specific cells example', color_palette = emb_palette, 
                            aligningCell = cellsExp[1], align_t = 0, cells = cellsExp, 
                            xrange = xrange, yrange = yrange)
fig_blot_cells
```
#### several lineages in each embryo
```{r}
lineExp <- c("ABalp", "ABara", "ABplpappp", "ABprpappp")
fig_blot_lines <- embryos|>plotBlotLine(title = 'lineages example', 
                            aligningCell = lineExp[1], align_t = 0, lineages = lineExp, 
                            color_palette = emb_palette,
                            xrange = xrange, yrange = yrange)
fig_blot_lines
```
####Syntax examples for extracting a single CD from such a list to allow applying 3D functions or tree drawing
```{r}
library(reticulate)

exEmbryo <- embryos[["CD"]][[1]]
exEmbryo <- totalRePosition(
  exEmbryo, time = time, indicatorP = "C", indicatorD = "Cxa", indicatorV = "MSxxp",
  xSize=0.08651, ySize=0.08651, zSize=0.5) #the xyz dimension is corrected during rotation
fig_3D_list <- drawEmbLine(
  exEmbryo, time 
##  xSize=0.08651, ySize=0.08651, zSize=0.5 #unprocessed data can have different x,y,z scales, specify here
  )
fig_3D_list[[1]]
```

####save one frame as a png, keeping color scale legend
```{r}
saveEmbImg(fig_3D_list[[1]], output_file="example_3D.png",keepLabels=TRUE)
```


##save a collection of frames that could be combined make a movie (in FIJI or other image analysis software)
```{r}
outdir <- "out/"
dir.create("out", recursive = TRUE, showWarnings = FALSE)

for(time in 1:150){
    if(time %% 10 == 0){
        message(time)
    }
    fig_3D_list <- drawEmbLine(
        exEmbryo, time, cellSize=15,
        xSize=1, ySize=1, zSize=1 #unprocessed data can have different x,y,z scales, specify here
    )
    saveEmbImg(fig_3D_list[[1]], output_file=paste0(outdir, "example_3D", time,".png"),keepLabels=TRUE, width = 800, height = 650, zoom=1.3)
}
```
